---
title: 'R Tutorial: Precipitation Graphs, Pt 2'
author: "Nathan Byers"
date: "Monday, July 17, 2014"
output:
  html_document:
    fig_height: 6
    fig_width: 9
    theme: readable
---

In <a href="http://rpubs.com/NateByers/precipRidem" target="_blank">part 1</a>
of this tutorial we tidied up some data from the `IDEMdata` package. In
this part of the tutorial we will use that data to create precipitation graphs. 
And in part 3 we will use functions and loops to automate the creation of these
graphs. 

# The data


We will be using the `wide.precip.df` data frame to create the precipitation
graphs. See <a href="http://rpubs.com/NateByers/precipRidem" target="_blank">part 1</a>
for instructions on how this data frame was created. Use the `date()` function
to load the data from the `IDEMdata` package.

```{r, eval=FALSE}
library(devtools)
install_github("InDEM/IDEMdata")
library(IDEMdata)
data(wide_precip)
```

```{r, echo=FALSE}
load("wide_precip_df.rdata")
```

# The graphs

For the graphs, we will first use the base `barplot()` function to display 
precipitation levels over time, with the chemistry sampling values overlayed as
points. Then we will also use the `lattice` and `ggplot2` packages to 
display the same data in alternative ways.

## Base graphics

For the base graphics plot, we first display the precipitation data as a bar plot. 
We also save the plot as a variable to retain information about its dimensions.

```{r, eval=FALSE}
op <- par(mar=c(5, 4, 4, 6) + 0.1)
precip <- unique(wide.precip.df[, c("date", "precipitation")])
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
```

```{r, echo=FALSE}
op <- par(mar=c(5, 4, 4, 6) + 0.1)
precip <- unique(wide.precip.df[, c("date", "precipitation")])
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
par(op)
```

We will be using this plot as a back drop for plotting the watershed sampling
data as points. The first plot will be percent saturation for station LMG-05-0009.

```{r}
station.1 <- wide.precip.df[wide.precip.df$station == "LMG-05-0009" &
                              !is.na(wide.precip.df$station), ]
```

In order to add points to a bar plot, we'll use the x locations that are found
in the `b.plot` variable. 

```{r, eval=FALSE}
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
bar.x <- b.plot[station.1$jday]

par(new = TRUE)
plot(x = bar.x, y = station.1$sat, xlab = "", ylab = "", pch = 16,
     ylim = c(0.5*min(station.1$sat, na.rm = TRUE), 2*max(station.1$sat, na.rm = TRUE)),
     axes = FALSE, col = "red")
axis(4, col = "red", col.axis = "red")
mtext("% Saturation", side = 4, line=3, cex = par("cex.lab"), col = "red")
title(main="Site 1")
par(op)  # reset graphing parameters
```

```{r, echo=FALSE}
op <- par(mar=c(5, 4, 4, 6) + 0.1)
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
bar.x <- b.plot[station.1$jday]

par(new = TRUE)
plot(x = bar.x, y = station.1$sat, xlab = "", ylab = "", pch = 16,
     ylim = c(0.5*min(station.1$sat, na.rm = TRUE), 2*max(station.1$sat, na.rm = TRUE)),
     axes = FALSE, col = "red")
axis(4, col = "red", col.axis = "red")
mtext("% Saturation", side = 4, line=3, cex = par("cex.lab"), col = "red")
title(main="Site 1")

```

Finally, we add a legend.

```{r, eval=FALSE}
legend("topright", # places a legend at the appropriate place 
       legend = c("Precip","% Sat"), # puts text in the legend 
       lty=c(1, 0), pch = c(NA, 19), # gives the legend appropriate symbols 
       col=c("black","red")) # appropriate colors
par(op)  # reset graphing parameters

```

```{r, echo=FALSE}
op <- par(mar=c(5, 4, 4, 6) + 0.1)
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
bar.x <- b.plot[station.1$jday]

par(new = TRUE)
plot(x = bar.x, y = station.1$sat, xlab = "", ylab = "", pch = 16,
     ylim = c(0.5*min(station.1$sat, na.rm = TRUE), 2*max(station.1$sat, na.rm = TRUE)), 
     axes = FALSE, col = "red")
axis(4, col = "red", col.axis = "red")
mtext("% Saturation", side = 4, line=3, cex = par("cex.lab"), col = "red")
title(main="Site 1")

legend("topright", # places a legend at the appropriate place 
       legend = c("Precip","% Sat"), # puts text in the legend 
       lty=c(1, 0), pch = c(NA, 19), # gives the legend appropriate symbols
       col=c("black","red")) # appropriate colors
par(op)  # reset graphing parameters
```


Now we'll make another graph, this time using E. coli data.

```{r, eval=FALSE}
op <- par(mar=c(5, 4, 4, 6) + 0.1)
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
bar.x <- b.plot[station.1$jday]

par(new = TRUE)
plot(x = bar.x, y = station.1$e_coli, xlab = "", ylab = "", pch = 16,
     ylim = c(0.5*min(station.1$e_coli, na.rm = TRUE), 2*max(station.1$e_coli, na.rm = TRUE)), 
     axes = FALSE, col = "red")
axis(4, col = "red", col.axis = "red")
abline(h=125, lty=2, col="red")
mtext("E. coli (MPN/100mL)", side = 4, line=3, cex = par("cex.lab"), col = "red")
title(main="Site 1")

legend("topright", # places a legend at the appropriate place 
       legend = c("Precip","E. coli", "WQS"), # puts text in the legend 
       lty=c(1, 0, 2), pch = c(NA, 19, NA), # gives the legend appropriate symbols 
       col=c("black","red", "red")) # appropriate colors
par(op)  # reset graphing parameters
```

```{r, echo=FALSE}
op <- par(mar=c(5, 4, 4, 6) + 0.1)
b.plot <- barplot(height = precip$precipitation, names.arg = precip$date,
                  xlab="Date", ylab="Precipitation (in)")
bar.x <- b.plot[station.1$jday]

par(new = TRUE)
plot(x = bar.x, y = station.1$e_coli, xlab = "", ylab = "", pch = 16,
     ylim = c(0.5*min(station.1$e_coli, na.rm = TRUE), 2*max(station.1$e_coli, na.rm = TRUE)), 
     axes = FALSE, col = "red")
axis(4, col = "red", col.axis = "red")
abline(h=125, lty=2, col="red")
mtext("E. coli (MPN/100mL)", side = 4, line=3, cex = par("cex.lab"), col = "red")
title(main="Site 1")

legend("topright", # places a legend at the appropriate place 
       legend = c("Precip","E. coli", "WQS"), # puts text in the legend 
       lty=c(1, 0, 2), pch = c(NA, 19, NA), # gives the legend appropriate symbols 
       col=c("black","red", "red")) # appropriate colors
par(op)  # reset graphing parameters
```

##lattice

`lattice` is another very popular graphics package. For an introduction and examples
see <a href="http://www.statmethods.net/advgraphs/trellis.html"
target="_blank">here</a> and <a href="http://www.isid.ac.in/~deepayan/R-tutorials/labs/04_lattice_lab.pdf" target="_blank">here</a>. We'll also use another package called 
`latticeExtra` which will make it easier to add the second y-axis to our package

Here's a first attempt at a `lattice` graph:

```{r, message=FALSE}
library(lattice)
library(latticeExtra)

rain <- xyplot(precipitation ~ date, data = wide.precip.df, type = "h",
               ylab = "Precipitation (in)")
sat <- xyplot(sat  ~ date, data = station.1, type = "p", ylab = "% Saturation", 
              ylim = c(0, 2*max(station.1$sat, na.rm = TRUE)))

doubleYScale(rain, sat, add.ylab2 = TRUE, under = TRUE,
             text = c( "Precipitation (in)", "% Saturation"), columns = 2, 
             type = c("l", "p"))

```

We can change the format of the plot to look like our base bar plot by using the 
`update()` function.

```{r}
update(trellis.last.object(),
       par.settings = simpleTheme(col = c("black", "red"), pch = c(NA, 16)))
```

The advantage of the `lattice` package is that it allows you to easily split up
the data by conditioning. We'll create some variables that will allow us to 
split up our data by time periods.

```{r}
# create a data set for one location in a format for lattice conditioning
lat.data <- merge(wide.precip.df[, c("date", "precipitation")], 
                  station.1[, c("date", "sat")], all = TRUE)
lat.data <- lat.data[!duplicated(lat.data), ]
lat.data.2013 <- lat.data[substring(as.character(lat.data$date), 1, 4) == "2013", ]


# create time variables
time <- as.POSIXlt(lat.data.2013$date)
month <- factor(months(time), levels = month.name[4:12])
mday <- time$mday
lat.data.2013 <- data.frame(lat.data.2013, month, mday)
```

Now we'll creat two lattice plots, save them as variables, and display them
together using the `doubleYScale()` function.

```{r}
rain <- xyplot(precipitation ~ mday | month, data = lat.data.2013, type = "h",
               ylab = "Precipitation (in)", layout = c(3, 3), as.table = TRUE, 
               xlab = "2013")
sat <- xyplot(sat  ~ mday | month, data = lat.data.2013, type = "p", 
              ylab = "% Saturation", ylim = c(0, 2*max(station.1$sat, na.rm = TRUE)))

doubley <- doubleYScale(rain, sat, add.ylab2 = TRUE, under = TRUE, 
                        text = c( "Precipitation (in)", "% Saturation"),
                        column = 2, type = c("l", "p"))
update(doubley,
       par.settings = simpleTheme(col = c("black", "red"), pch = c(NA, 16)))

```

## ggplot2

The `ggplot2` package is also useful for displaying relationships between variables
by conditioning. However, it's not possible to add two y-axes to the same plot 
using `ggplot2`. So we will create two cumulative variables that will give us
information about rainfall in days prior to the date of the field samples. One 
variable will be the sum of rainfall on the same day and one day prior to the
sample. Another variable will be the sum of the same day, one day and 2 days
prior to the sample.

```{r}
precip_2day <- lat.data$precipitation   +
  c(0, lat.data$precipitation)[1:length(lat.data$precipitation)]
precip_3day <- lat.data$precipitation   +
  c(0, lat.data$precipitation)[1:length(lat.data$precipitation)] +
  c(0, 0, lat.data$precipitation)[1:length(lat.data$precipitation)]
gg.data <- data.frame(lat.data, precip_2day, precip_3day)
```

Now we reshape this long data frame into a wide format and plot it.

```{r, message=FALSE, warning=FALSE}
library(reshape2)
long.gg.data <- melt(gg.data, id.vars = c("date", "sat"))

library(ggplot2)
ggplot(long.gg.data, aes(x = date, y = sat, size = value)) + facet_grid(variable ~ .) +
  geom_point(color = "steelblue") + scale_size_continuous(range = c(3,8)) + 
  ylab("% Saturation") + xlab("Date") + labs(size = "Precip. (in)") +
  theme(strip.background = element_rect(fill="lemonchiffon1")) 
  
```



...
-------------------------------------------------------------------------------


### R Markdown

This tutorial was created using RStudio's <a href="http://rmarkdown.rstudio.com/"
target="_blank">R Markdown</a>. The code can be found on <a href="https://github.com/NateByers/precip2Ridem"
target="_blank">GitHub</a>.